% NLP Project 2.2 - English-Italian Machine Translation
% ACM proceedings template (sigconf).
%
% 'nonacm' drops the copyright block and makes CCS/keywords optional,
% and printacmref=false removes the ACM reference-format box.
% Together these save roughly one full column - which we need for the 4-page limit.
\documentclass[sigconf,nonacm]{acmart}
\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

\usepackage{booktabs}
\usepackage{graphicx}

\begin{document}

\title{English--Italian Neural Machine Translation on Europarl}

\author{Vasileios Lalos}
\affiliation{\institution{TU Berlin}\country{}}
\author{Riccardo Tella}
\affiliation{\institution{TU Berlin}\country{}}

\begin{abstract}
We build and compare recurrent encoder--decoder models for English--Italian
translation on the Europarl v7 corpus. Starting from a vanilla RNN baseline we
show that a fixed context vector is the dominant failure mode, and address it
progressively with context feeding, LSTM units, pre-trained embeddings and
finally attention. We report BLEU and METEOR for every variant, compare both
translation directions, contrast word- and character-level models, and analyse
how performance degrades with sentence length. % TODO: add one sentence with the headline result once Tasks 4-5 are in.
\end{abstract}

\maketitle

\section{Introduction}
% Keep this SHORT - the writing suggestions explicitly warn against long
% background sections. Three or four sentences is enough.
Machine translation maps a sentence in a source language onto a sentence in a
target language. Unlike classification there is no single correct output, so
translation quality is measured by comparing system output against reference
translations. We study English--Italian translation on the Europarl parallel
corpus, focusing on how architectural and representational choices affect
model behaviour rather than on maximising a single score.

\section{Data and Exploratory Analysis}
\label{sec:data}

We use the Europarl v7 English--Italian parallel corpus, which contains
1,909,115 line-aligned sentence pairs. Table~\ref{tab:corpus} summarises it.

\begin{table}[t]
\caption{Europarl v7 English--Italian corpus statistics (full corpus).}
\label{tab:corpus}
\begin{tabular}{lrr}
\toprule
 & \textbf{English} & \textbf{Italian} \\
\midrule
Sentence pairs      & 1,909,115 & 1,909,115 \\
Total tokens        & 49.7\,M   & 48.0\,M   \\
Vocabulary (types)  & 275,537   & \textbf{426,025} \\
Mean length (tokens)& 26.05     & 25.13 \\
Median length       & 23        & 23 \\
Max length          & 668       & 558 \\
Empty lines         & 3,257     & 4,985 \\
\bottomrule
\end{tabular}
\end{table}

The most informative asymmetry is lexical rather than positional. Italian uses
slightly \emph{fewer} tokens than English on average (25.13 vs.\ 26.05) yet
requires 55\% more distinct word types (426,025 vs.\ 275,537). This reflects
Italian's richer inflectional morphology: a single verb lemma surfaces as many
conjugated forms, each a separate entry for a word-level model. Two practical
consequences follow. First, under a fixed vocabulary cap the Italian side
suffers more out-of-vocabulary loss, so generating Italian is harder than
generating English -- a prediction we test in
Section~\ref{sec:direction}. Second, it motivates the character-level model of
Section~\ref{sec:char}, which avoids the vocabulary explosion entirely.

We also note that the common assumption that Romance translations are longer
than their Germanic sources does not hold at the token level here; it holds in
characters, not words. Sentence lengths are strongly correlated across the two
languages ($r = 0.946$), which makes length a clean axis for the error analysis
in Section~\ref{sec:length}.

Because the full corpus is too large to train on, we randomly sample 10\% of
the aligned pairs (190,912 pairs, seed 42). Sampling is performed over
\emph{indices} so that pair alignment is preserved, and the drawn indices are
stored so the sample is exactly reproducible.

\section{Preprocessing}
\label{sec:prep}

Preprocessing removes a sentence pair whenever \emph{either} side is unusable,
which preserves alignment. We lowercase, normalise repeated whitespace, strip
leading and trailing whitespace, drop empty pairs, and drop pairs containing
XML-like lines. Of 190,912 pairs, 802 (0.42\%) were removed, all of them empty;
this release of Europarl contains no XML-tag lines, so that filter was a
no-op safety check.

We deliberately \emph{retain} punctuation, numbers and stopwords, and apply
neither stemming nor lemmatisation. This is the opposite of the choice one
makes for text classification, and for a specific reason: in translation every
token must be reproduced in the output. Function words carry grammatical
information the decoder needs to generate, and stemming would destroy exactly
the inflectional endings that Italian generation depends on.

\section{Neural Machine Translation}
\label{sec:nmt}

% TODO (Riccardo): architecture description + justification.
% Cover: embedding -> encoder RNN/LSTM -> fixed context -> decoder, teacher
% forcing, greedy decoding, cross-entropy ignoring padding, Adam, grad clipping.

Sentence pairs longer than 20 tokens on either side are discarded, leaving
70,914 pairs, split 70/10/20 into training (49,641), validation (7,091) and
test (14,182) sets with seed 42. Vocabularies of 10,000 types per language are
built on the training split only. All models are evaluated with corpus BLEU and
mean sentence-level METEOR, both on a 0--100 scale.

\subsection{From vanilla RNN to LSTM}
% TODO (Riccardo): the diagnostic narrative. This is the most valuable part of
% the report - the failures matter more than the final number.
The vanilla RNN encoder--decoder collapsed to a single generic output such as
\texttt{la <unk>}. Training longer did not help and eventually overfitted.
Table~\ref{tab:tuning} tracks the fixes.

\begin{table}[t]
\caption{Model development, validation set. Best per column in bold.}
\label{tab:tuning}
\begin{tabular}{lrrr}
\toprule
\textbf{Model} & \textbf{Val.\ loss} & \textbf{BLEU} & \textbf{METEOR} \\
\midrule
Context-fed RNN        & 4.208 & 1.84 & 11.11 \\
LSTM, hidden 128       & 3.892 & 3.25 & 16.53 \\
LSTM, hidden 256       & 3.782 & 3.92 & 18.23 \\
LSTM 256 + Word2Vec    & \textbf{3.560} & \textbf{4.17} & \textbf{19.23} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Embeddings}
% TODO: extend if FastText is added.
Word2Vec (CBOW) embeddings trained on the training split only, then fine-tuned,
improve every metric over random initialisation.

\subsection{Translation direction}
\label{sec:direction}
% TODO (Riccardo): IT->EN results and comparison with EN->IT.
% Link back to the morphology finding in Section 2.

\subsection{Character-level model}
\label{sec:char}
% TODO (Riccardo): char model vs word model.

\subsection{Effect of sentence length}
\label{sec:length}
% TODO: BLEU bucketed by source length. This directly answers the brief's
% question "does the length of text impact the performance of the model?".

\section{Attention}
\label{sec:attn}
% TODO (Vasilis): Task 4.
% - describe the attention mechanism used (Bahdanau or Luong) and why
% - table: with vs without attention, same data and hyperparameters
% - attention heatmap figure for one sample sentence
% - tie back: attention removes the fixed-context bottleneck, so the gain
%   should be largest on LONG sentences - check that against Section 4.5.

\section{Pivot Translation}
\label{sec:pivot}
% TODO (Vasilis): Task 5. Italian -> English -> Swedish.
% Discuss error propagation: errors in IT->EN are inherited by EN->SV.

\section{Conclusion}
% TODO: 3-4 sentences. What changed the results most, what we would do next.

\end{document}
